---
title: "Evaluating KOMPUTE - Open Field Data"
author: "Coby Warkentin and Donghyung Lee"
date: "`r Sys.Date()`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Load packages
```{r load_packages, echo=TRUE, message=FALSE}
rm(list=ls())
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

library(data.table)
library(dplyr)
library(reshape2)
library(ggplot2)
library(tidyr) #spread
library(RColorBrewer)
library(circlize)
library(ComplexHeatmap)
```

## Preparing control phenotype data

### Importing Open Field Control Phenotype Dataset
```{r load_data, echo=TRUE}
OF.data <- readRDS("data/OF.data.rds")
dim(OF.data)
```

### Visualizing measured phenotypes via a heatmap 
The heatmap below presents a visualization of the phenotypic measurements taken for each control mouse. The columns represent individual mice, while the rows correspond to the distinct phenotypes measured. 
```{r heatmap_pheno, fig.width=12, fig.height=14}
mtest <- table(OF.data$proc_param_name_stable_id, OF.data$biological_sample_id)
mtest <-as.data.frame.matrix(mtest)
dim(mtest)
if(FALSE){
nmax <-max(mtest)
library(circlize)
col_fun = colorRamp2(c(0, nmax), c("white", "red"))
col_fun(seq(0, nmax))
ht = Heatmap(as.matrix(mtest), cluster_rows = FALSE, cluster_columns = FALSE, show_column_names = F, col = col_fun,
             row_names_gp = gpar(fontsize = 8), name="Count")
draw(ht)
}
```

### Exclude phenotypes with fewer than 15,000 observations
To maintain data quality and robustness, we will discard any phenotypes that have fewer than 15,000 recorded observations.
```{r}
mtest <- table(OF.data$proc_param_name, OF.data$biological_sample_id)
dim(mtest)
#head(mtest[,1:10])
mtest0 <- mtest>0
#head(mtest0[,1:10])
rowSums(mtest0)
rmv.pheno.list <- rownames(mtest)[rowSums(mtest0)<15000]
#rmv.pheno.list
dim(OF.data)
OF.data <- OF.data %>% filter(!(proc_param_name %in% rmv.pheno.list))
dim(OF.data)
# number of phenotypes left
length(unique(OF.data$proc_param_name))
```

### Remove samples with fewer than 10 measured phenotypes
```{r rmv_samp, echo=TRUE}
mtest <- table(OF.data$proc_param_name, OF.data$biological_sample_id)
dim(mtest)
head(mtest[,1:10])
mtest0 <- mtest>0
head(mtest0[,1:10])
summary(colSums(mtest0))
rmv.sample.list <- colnames(mtest)[colSums(mtest0)<10]
length(rmv.sample.list)
dim(OF.data)
OF.data <- OF.data %>% filter(!(biological_sample_id %in% rmv.sample.list))
dim(OF.data)
# number of observations to use
length(unique(OF.data$biological_sample_id))
```

### Heapmap of filtered phenotypes
```{r}
if(FALSE){
mtest <- table(OF.data$proc_param_name, OF.data$biological_sample_id)
dim(mtest)
mtest <-as.data.frame.matrix(mtest)
nmax <-max(mtest)
library(circlize)
col_fun = colorRamp2(c(0, nmax), c("white", "red"))
col_fun(seq(0, nmax))
pdf("~/Google Drive Miami/Miami_IMPC/output/measured_phenotypes_controls_after_filtering_OF.pdf", width = 10, height = 3)
ht = Heatmap(as.matrix(mtest), cluster_rows = FALSE, cluster_columns = FALSE, show_column_names = F, col = col_fun,
             row_names_gp = gpar(fontsize = 7), name="Count")
draw(ht)
dev.off()
}
```

### Reforatting the dataset (long to wide)

We restructure our data from a long format to a wide one for further processing and analysis.
```{r reshape_pheno}
OF.mat <- OF.data %>% 
  dplyr::select(biological_sample_id, proc_param_name, data_point, sex, phenotyping_center, strain_name) %>% 
  ##consider weight or age in weeks
  arrange(biological_sample_id) %>%
  distinct(biological_sample_id, proc_param_name, .keep_all=TRUE) %>% ## remove duplicates, maybe mean() is better.
  spread(proc_param_name, data_point) %>%
  tibble::column_to_rownames(var="biological_sample_id")
head(OF.mat)
dim(OF.mat)
summary(colSums(is.na(OF.mat[,-1:-3])))
```

### Visualizing phenotype distributions 
```{r pheno_dist,  fig.width=10, fig.height=5, message=FALSE, warning=FALSE}
ggplot(melt(OF.mat), aes(x=value)) + 
  geom_histogram() + 
  facet_wrap(~variable, scales="free", ncol=5)+
  theme(strip.text.x = element_text(size = 6))
```


### Rank Z transformation

In this step, we conduct a rank Z transformation on the phenotype data to ensure that the data is normally distributed
```{r rankZ_trans, fig.width=10, fig.height=5, message=FALSE, warning=FALSE}
library(RNOmni)
OF.mat.rank <- OF.mat
dim(OF.mat.rank)
OF.mat.rank <- OF.mat.rank[complete.cases(OF.mat.rank),]
dim(OF.mat.rank)

dim(OF.mat)
OF.mat <- OF.mat[complete.cases(OF.mat),]
dim(OF.mat)

OF.mat.rank <- cbind(OF.mat.rank[,1:3], apply(OF.mat.rank[,-1:-3], 2, RankNorm))
ggplot(melt(OF.mat.rank), aes(x=value)) + 
  geom_histogram() + 
  facet_wrap(~variable, scales="free", ncol=5)+
  theme(strip.text.x = element_text(size = 6))
```

### Conducting Principal Variance Component Analysis (PVCA)

In this step, we apply Principal Variance Component Analysis (PVCA) on the phenotype matrix data. PVCA is an approach that combines Principal Component Analysis (PCA) and Variance Component Analysis to quantify the proportion of total variance in the data attributed to each important covariate, in this case 'sex' and 'phenotyping_center'.

First, we prepare our metadata which includes our chosen covariates. Any character variables in the metadata are then converted to factors. To avoid potential confounding, we check for associations between our covariates and drop 'strain_name' due to its strong association with 'phenotyping_center'.

Next, we run PVCA on randomly chosen subsets of our phenotype data (for computational efficiency). Finally, we compute the average effect size across all random samples and visualize the results in a PVCA plot.

```{r PVCA, fig.width=6, fig.height=5, message=FALSE, warning=FALSE}
source("code/PVCA.R")

meta <- OF.mat.rank[,1:3] ## examining covariates sex, phenotyping_center, and strain_name
head(meta)
dim(meta)
summary(meta) # variables are currently characters
meta[sapply(meta, is.character)] <- lapply(meta[sapply(meta, is.character)], as.factor)
summary(meta) # now all variables are converted to factors

chisq.test(meta[,1],meta[,2])
chisq.test(meta[,2],meta[,3]) 

meta<-meta[,-3] # phenotyping_center and strain_name strongly associated which could cause confounding in the PVCA analysis, so we drop 'strain_name'.

G <- t(OF.mat.rank[,-1:-3])  # preparing the phenotype matrix data

set.seed(09302021)

# Perform PVCA for 10 random samples of size 1000 (more computationally efficient)
pvca.res <- matrix(nrow=10, ncol=3)
for (i in 1:10){
  sample <- sample(1:ncol(G), 1000, replace=FALSE)
  pvca.res[i,] <- PVCA(G[,sample], meta[sample,], threshold=0.6, inter=FALSE)
}

# Compute average effect size across the 10 random samples
pvca.means <- colMeans(pvca.res)
names(pvca.means) <- c(colnames(meta), "resid")

# Create PVCA plot
pvca.plot <- PlotPVCA(pvca.means, "PVCA of Phenotype Matrix Data")
pvca.plot

png(file="docs/figure/figures.Rmd/pvca_OF_1_v16.png", width=600, height=350)
pvca.plot
dev.off()

```

### Batch effect removal using ComBat
We remove batch effects (the center effect) in the phenotype data set by using the ComBat method.
```{r combat}
library(sva)
combat_komp = ComBat(dat=G, batch=meta$phenotyping_center, par.prior=TRUE, prior.plots=TRUE, mod=NULL)

combat_komp[1:5,1:5]
G[1:5,1:5] # for comparison, combat_komp is same form and same dimensions as G
```

### PVCA on ComBat residuals 
After using ComBat to account for batch effects, we perform a PVCA on the residuals. We expect to observe a significantly reduced effect from the phenotyping centers.
```{r PVCA_resid, fig.width=6, fig.height=5, message=FALSE, warning=FALSE}
set.seed(09302021)
# Perform PVCA for 10 random samples (more computationally efficient)
pvca.res.nobatch <- matrix(nrow=10, ncol=3)
for (i in 1:10){
  sample <- sample(1:ncol(combat_komp), 1000, replace=FALSE)
  pvca.res.nobatch[i,] <- PVCA(combat_komp[,sample], meta[sample,], threshold=0.6, inter=FALSE)
}

# Compute average effect size across samples
pvca.means.nobatch <- colMeans(pvca.res.nobatch)
names(pvca.means.nobatch) <- c(colnames(meta), "resid")

# Generate PVCA plot
pvca.plot.nobatch <- PlotPVCA(pvca.means.nobatch, "PVCA of Phenotype Matrix Data with Reduced Batch Effect")
pvca.plot.nobatch

if(FALSE){
png(file="docs/figure/figures.Rmd/pvca_OF_2_v16.png", width=600, height=350)
pvca.plot.nobatch
dev.off()
}
```


### Computing phenotypic correlations
We compute the phenotype correlations using different methods and compare them.
```{r, fig.width=6, fig.height=3}
OF.cor.rank <- cor(OF.mat.rank[,-1:-3], use="pairwise.complete.obs") # pearson correlation coefficient
OF.cor <- cor(OF.mat[,-1:-3], use="pairwise.complete.obs", method="spearman") # spearman
OF.cor.combat <- cor(t(combat_komp), use="pairwise.complete.obs")
pheno.list <- rownames(OF.cor)

ht1 = Heatmap(OF.cor, show_column_names = F, row_names_gp = gpar(fontsize = 9), name="Spearm. Corr.")
draw(ht1)
ht2 = Heatmap(OF.cor.rank, show_column_names = F, row_names_gp = gpar(fontsize = 9), name="Corr. RankZ")
draw(ht2)
ht3 = Heatmap(OF.cor.combat, show_column_names = F, row_names_gp = gpar(fontsize = 9), name="Corr. ComBat")
draw(ht3)
```


## Preparation of IMPC summary statistics data

### Loading Open Field summary stat (IMPCv16)
```{r KOMP_stats}
OF.stat <- readRDS("data/OF.stat.v16.rds")
dim(OF.stat)
table(OF.stat$parameter_name, OF.stat$procedure_name)

length(unique(OF.stat$marker_symbol)) #3362
length(unique(OF.stat$allele_symbol)) #3412
length(unique(OF.stat$proc_param_name)) #15, number of phenotypes in association statistics data set
length(unique(OF.data$proc_param_name)) #15, number of phenotypes in final control data

pheno.list.stat <- unique(OF.stat$proc_param_name)
pheno.list.ctrl <- unique(OF.data$proc_param_name)
sum(pheno.list.stat %in% pheno.list.ctrl)
sum(pheno.list.ctrl %in% pheno.list.stat)

# Identifying common phenotypes between statistics and control data
common.pheno.list <- sort(intersect(pheno.list.ctrl, pheno.list.stat))
common.pheno.list
length(common.pheno.list) # 14 - each data set had one phenotype not present in the other

# Filtering summary statistics to contain only common phenotypes
dim(OF.stat)
OF.stat <- OF.stat %>% filter(proc_param_name %in% common.pheno.list)
dim(OF.stat)
length(unique(OF.stat$proc_param_name))
```

### Visualizing gene-phenotype pair duplicates 
```{r gene_pheno_dup, fig.width=6, fig.height=2}
mtest <- table(OF.stat$proc_param_name, OF.stat$marker_symbol)
mtest <-as.data.frame.matrix(mtest)
nmax <-max(mtest)
col_fun = colorRamp2(c(0, nmax), c("white", "red"))
col_fun(seq(0, nmax))
ht = Heatmap(as.matrix(mtest), cluster_rows = FALSE, cluster_columns = FALSE, show_column_names = F, col = col_fun,
             row_names_gp = gpar(fontsize = 8), name="Count")
draw(ht)
```

### Consolidating muliple z-scores of a gene-phenotype pair using Stouffer's Method
```{r Stouffer}
## sum(z-score)/sqrt(# of zscore)
sumz <- function(z){ sum(z)/sqrt(length(z)) }
OF.z = OF.stat %>%
  dplyr::select(marker_symbol, proc_param_name, z_score) %>%
  na.omit() %>%
  group_by(marker_symbol, proc_param_name) %>% 
  summarize(zscore = sumz(z_score)) ## combine z-scores
dim(OF.z)

```

### Generating Z-score matrix (reformatting)
```{r}
# Function to convert NaN to NA
nan2na <- function(df){ 
  out <- data.frame(sapply(df, function(x) ifelse(is.nan(x), NA, x))) 
  colnames(out) <- colnames(df)
  out
}

# Converting the long format of z-scores to a wide format matrix
OF.zmat = dcast(OF.z, marker_symbol ~ proc_param_name, value.var = "zscore", 
             fun.aggregate = mean) %>% tibble::column_to_rownames(var="marker_symbol")
OF.zmat = nan2na(OF.zmat) #convert nan to na
dim(OF.zmat)

dim(OF.zmat)
saveRDS(OF.zmat, file = "data/OF.zmat.v16.rds")
```

### Visualization of Phenotype-Gene Coverage
The heatmap illustrates tested (red) and untested (white) gene-phenotype pairs.
```{r, fig.width=6, fig.height=2.5}
# Generate a matrix indicating where z-scores are present
id.mat <- 1*(!is.na(OF.zmat)) # multiply 1 to make this matrix numeric
nrow(as.data.frame(colSums(id.mat)))
dim(id.mat)

ht = Heatmap(t(id.mat), 
             cluster_rows = T, clustering_distance_rows ="binary",
             cluster_columns = T, clustering_distance_columns = "binary",
             show_row_dend = F, show_column_dend = F,  # do not show dendrogram
             show_column_names = F, col = c("white","red"),
             row_names_gp = gpar(fontsize = 10), name="Missing")
draw(ht)
```

### Distribution of Z-Scores Across Phenotypes
The histogram presents the distribution of association Z-scores for each phenotype.
```{r, fig.width=10, fig.height=5, message=FALSE, warning=FALSE}
ggplot(melt(OF.zmat), aes(x=value)) + 
  geom_histogram() + 
  facet_wrap(~variable, scales="free", ncol=5)+
  theme(strip.text.x = element_text(size = 6))
```


### Estimation of Genetic Correlation Matrix Using Z-Scores

Here, we estimate the genetic correlations between phenotypes utilizing the association Z-score matrix.
```{r, fig.width=7, fig.height=3}
# Select common phenotypes
OF.zmat <- OF.zmat[,common.pheno.list]
dim(OF.zmat)

# Compute genetic correlations
OF.zcor = cor(OF.zmat, use="pairwise.complete.obs")

# Generate heatmap of the correlation matrix
ht = Heatmap(OF.zcor, cluster_rows = T, cluster_columns = T, show_column_names = F, #col = col_fun,
             row_names_gp = gpar(fontsize = 10),
             name="Genetic Corr (Z-score)"
             )
draw(ht)
```


### Comparison of Phenotypic Correlation and Genetic Correlation Among Phenotypes

We will compare the correlation matrix obtained from control mice phenotype data and the genetic correlation matrix estimated using association Z-scores. As depicted below, both correlation heatmaps show similar correlation patterns.
```{r comp_pheno_corr_gene_corr, fig.width=12, fig.height=2.6}

OF.cor.rank.fig <- OF.cor.rank[common.pheno.list,common.pheno.list]
OF.cor.fig <- OF.cor[common.pheno.list,common.pheno.list]
OF.cor.combat.fig <- OF.cor.combat[common.pheno.list, common.pheno.list]
OF.zcor.fig <- OF.zcor


ht = Heatmap(OF.cor.rank.fig, cluster_rows = TRUE, cluster_columns = TRUE, show_column_names = F, #col = col_fun,
              show_row_dend = F, show_column_dend = F,  # do not show dendrogram
             row_names_gp = gpar(fontsize = 8), column_title="Phenotype Corr (RankZ, Pearson)", column_title_gp = gpar(fontsize = 8),
             name="Corr")
pheno.order <- row_order(ht)
#draw(ht)

OF.cor.rank.fig <- OF.cor.rank.fig[pheno.order,pheno.order]
ht1 = Heatmap(OF.cor.rank.fig, cluster_rows = FALSE, cluster_columns = FALSE, show_column_names = F, #col = col_fun,
              show_row_dend = F, show_column_dend = F,  # do not show dendrogram
             row_names_gp = gpar(fontsize = 8), column_title="Phenotype Corr (RankZ, Pearson)", column_title_gp = gpar(fontsize = 8),
             name="Corr")
OF.cor.fig <- OF.cor.fig[pheno.order,pheno.order]  
ht2 = Heatmap(OF.cor.fig, cluster_rows = FALSE, cluster_columns = FALSE, show_column_names = F, #col = col_fun,
             row_names_gp = gpar(fontsize = 8), column_title="Phenotype Corr (Spearman)", column_title_gp = gpar(fontsize = 8),
             name="Corr")
OF.cor.combat.fig <- OF.cor.combat.fig[pheno.order,pheno.order]  
ht3 = Heatmap(OF.cor.combat.fig, cluster_rows = FALSE, cluster_columns = FALSE, show_column_names = F, #col = col_fun,
             row_names_gp = gpar(fontsize = 8), column_title="Phenotype Corr (Combat, Pearson)", column_title_gp = gpar(fontsize = 8),
             name="Corr")
OF.zcor.fig <- OF.zcor.fig[pheno.order,pheno.order]
ht4 = Heatmap(OF.zcor.fig, cluster_rows = FALSE, cluster_columns = FALSE, show_column_names = F, #col = col_fun,
             row_names_gp = gpar(fontsize = 8), column_title="Genetic Corr  (Pearson)", column_title_gp = gpar(fontsize = 8),
             name="Corr"
             )
draw(ht1+ht2+ht3+ht4)

if(FALSE){
png(file="docs/figure/figures.Rmd/cors_OF.png", width=800, height=250)
draw(ht1+ht2+ht3+ht4)
dev.off()
}
```

### Correlation Analysis Between Genetic Correlation Matrices Using Mantel's Test

To evaluate the correlation between different genetic correlation matrices, we apply Mantel's test, which measures the correlation between two distance matrices.
```{r}
####################
# Use Mantel test 
# https://stats.idre.ucla.edu/r/faq/how-can-i-perform-a-mantel-test-in-r/
# install.packages("ade4")
library(ade4)

# Function to extract upper triangular elements of a matrix
to.upper<-function(X) X[upper.tri(X,diag=FALSE)]

a1 <- to.upper(OF.cor.fig)
a2 <- to.upper(OF.cor.rank.fig)
a3 <- to.upper(OF.cor.combat.fig)
a4 <- to.upper(OF.zcor.fig)

plot(a4, a1)
plot(a4, a2)
plot(a4, a3)

mantel.rtest(as.dist(1-OF.cor.fig), as.dist(1-OF.zcor.fig), nrepet = 9999) #nrepet = number of permutations
mantel.rtest(as.dist(1-OF.cor.rank.fig), as.dist(1-OF.zcor.fig), nrepet = 9999)
mantel.rtest(as.dist(1-OF.cor.combat.fig), as.dist(1-OF.zcor.fig), nrepet = 9999)
```


## Evaluating the KOMPUTE Imputation Algorithm

### Initializing the KOMPUTE Package
```{r kompute_load}
# Check if KOMPUTE is installed, if not, install it from GitHub using devtools
if(!"kompute" %in% rownames(installed.packages())){
  library(devtools)
  devtools::install_github("dleelab/kompute")
}
library(kompute)
```


### Simulation study - Comparison of imputed vs measured z-score values

In this section, we conduct a simulation study to compare the performance of the KOMPUTE method with the measured gene-phenotype association z-scores. We randomly select some of these measured z-scores, mask them, and then use the KOMPUTE method to impute them. We then compare the imputed z-scores with the measured ones.
```{r}
zmat <-t(OF.zmat) 
dim(zmat)

# filter genes with less than 1 missing data point (na)
zmat0 <- is.na(zmat)
num.na<-colSums(zmat0)
summary(num.na)

dim(zmat)
dim(zmat[,num.na<1])
dim(zmat[,num.na<5])
dim(zmat[,num.na<10])

# filter genes with less than 1 missing data point (na)
zmat <- zmat[,num.na<1]
dim(zmat)

# Set correlation method for phenotypes
#pheno.cor <- OF.cor.fig
#pheno.cor <- OF.cor.rank.fig
pheno.cor <- OF.cor.combat.fig
#pheno.cor <- OF.zcor.fig

#save phenotype correlation matrix 
saveRDS(pheno.cor, file = "data/OF.pheno.cor.v16.rds")

zmat <- zmat[rownames(pheno.cor),,drop=FALSE]
rownames(zmat)
rownames(pheno.cor)
colnames(pheno.cor)
npheno <- nrow(zmat)

# calculate the percentage of missing Z-scores in the original data 
100*sum(is.na(zmat))/(nrow(zmat)*ncol(zmat)) # 0%
nimp <- 1000 # # of missing/imputed Z-scores
set.seed(1111)

# find index of all measured zscores
all.i <- 1:(nrow(zmat)*ncol(zmat))
measured <- as.vector(!is.na(as.matrix(zmat)))
measured.i <- all.i[measured]

# mask 2000 measured z-scores
mask.i <- sort(sample(measured.i, nimp))
org.z = as.matrix(zmat)[mask.i]
zvec <- as.vector(as.matrix(zmat))
zvec[mask.i] <- NA
zmat.imp <- matrix(zvec, nrow=npheno)
rownames(zmat.imp) <- rownames(zmat)
```

### Run KOMPUTE method
```{r KOMPUTE_res, fig.width=5, fig.height=4}
kompute.res <- kompute(zmat.imp, pheno.cor, 0.01)

# Compare measured vs imputed z-scores
length(org.z)
imp.z <- as.matrix(kompute.res$zmat)[mask.i]
imp.info <- as.matrix(kompute.res$infomat)[mask.i]  
plot(imp.z, org.z)

# Create a dataframe with the original and imputed z-scores and the information of imputed z-scores
imp <- data.frame(org.z=org.z, imp.z=imp.z, info=imp.info)
dim(imp)
imp <- imp[complete.cases(imp),]
imp <- subset(imp, info>=0 & info <= 1)
dim(imp)
cor.val <- round(cor(imp$imp.z, imp$org.z), digits=3)
cor.val
plot(imp$imp.z, imp$org.z)

# Set a cutoff for information content and filter the data accordingly
info.cutoff <- 0.8
imp.sub <- subset(imp, info>info.cutoff)
dim(imp.sub)
summary(imp.sub$imp.z)
summary(imp.sub$info)

cor.val <- round(cor(imp.sub$imp.z, imp.sub$org.z), digits=3)
cor.val
g <- ggplot(imp.sub, aes(x=imp.z, y=org.z, col=info)) +
    geom_point() +
    labs(title=paste0("IMPC Behavior Data (OF), Info>", info.cutoff, ", Cor=",cor.val),
      x="Imputed Z-scores", y = "Measured Z-scores", col="Info") +
    theme_minimal()
g

# save plot
png(file="docs/figure/figures.Rmd/sim_results_OF_v16.png", width=600, height=350)
g
dev.off()

# Part 3 of  Figure 2
fig2.3 <- ggplot(imp.sub, aes(x=imp.z, y=org.z, col=info)) +
  geom_point() +
  labs(title="Open Field",
       x="Imputed Z-scores", y = "", col="Info") +
  scale_x_continuous(limits=c(-9,9), breaks=c(seq(-9,9,3)), minor_breaks = NULL) +
  scale_y_continuous(limits=c(-9,9), breaks=c(seq(-9,9,3))) +
  scale_color_gradient(limits=c(0.8,1), low="#98cdf9", high="#084b82") +
  theme_bw() +
  #theme(legend.position="none", plot.title=element_text(hjust=0.5))
  theme(plot.title=element_text(hjust=0.5))
save(fig2.3, file="docs/figure/figures.Rmd/sim_OF_v16.rdata")
```
