---
title: "Simulation Study: Performance Comparision of KOMPUTE and SVD Matrix Completion"
author: "Coby Warkentin and Donghyung Lee"
date: "`r Sys.Date()`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


## Import the necessary packages
```{r}
library(MASS)
library(kompute)
library(kableExtra)
```

## Import source functions for the analysis
```{r}
source("code/svd_impute.R")
source("code/sim_function.R")
```

## Load the phenotypic correlation matrix 

Load the phenotypic correlation matrix specific to the Body Composition domain and set initial parameters
```{r}
pheno.cor <- readRDS("data/BC_pheno_cor.RDS")
n.genes <- 10000 # number of genes
seed <- 05152022 # set seed
```


## Simulations

### Matrix Completion

Simulate an association Z-score matrix with 10,000 genes and 9 phenotypes at varying levels of masking and impute the masked z-scores using the Matrix Completion method
```{r}
mc.res20 <- simulation(n.genes=n.genes, pheno.cor=pheno.cor, mask.prop=.2, seed=seed)
mc.res40 <- simulation(n.genes=n.genes, pheno.cor=pheno.cor, mask.prop=.4, seed=seed)
mc.res60 <- simulation(n.genes=n.genes, pheno.cor=pheno.cor, mask.prop=.6, seed=seed)
```


### KOMPUTE

Perform identical simulations as above, but impute the masked z-scores using the KOMPUTE method. Here, we include all imputed z-scores when computing the correlation coefficients between imputed and origianl z-scores
```{r}
komp.res20 <- simulation(n.genes=n.genes, pheno.cor=pheno.cor, mask.prop=.2, method="kompute", seed=seed)
komp.res40 <- simulation(n.genes=n.genes, pheno.cor=pheno.cor, mask.prop=.4, method="kompute", seed=seed)
komp.res60 <- simulation(n.genes=n.genes, pheno.cor=pheno.cor, mask.prop=.6, method="kompute", seed=seed)
```


### KOMPUTE with info > 0.8

Perform identical simulations and use the same method (KOMPUTE), however, we only utilize imputed z-scores that have an imputation information value greater than 0.8.
```{r}
komp.info.res20 <- simulation(n.genes=n.genes, pheno.cor=pheno.cor, mask.prop=.2, method="kompute", info.cutoff = 0.8, seed=seed)
komp.info.res40 <- simulation(n.genes=n.genes, pheno.cor=pheno.cor, mask.prop=.4, method="kompute", info.cutoff = 0.8, seed=seed)
komp.info.res60 <- simulation(n.genes=n.genes, pheno.cor=pheno.cor, mask.prop=.6, method="kompute", info.cutoff = 0.8, seed=seed)
```


## Results: Table

Complie a table of correlation coefficients from all the simulations conducted
```{r}
cor.table <- data.frame(c(mc.res20$cor, komp.res20$cor, komp.info.res20$cor),
                        c(mc.res40$cor, komp.res40$cor, komp.info.res40$cor),
                        c(mc.res60$cor, komp.res60$cor, komp.info.res60$cor))
colnames(cor.table) <- c("20% Removed", "40% Removed", "60% Removed")
rownames(cor.table) <- c("MC", "KPT", "KPT w/ info > 0.8")


cor.table %>%
  kbl(caption = "Pearson correlation coefficients between the original and imputed Z-scores") %>%
  kable_classic(full_width = F, html_font = "Cambria")
```


## Results: Figure

```{r, fig.height=15, fig.width=15}
library(gridExtra)

p1 <- mc.res20$plot
p2 <- mc.res40$plot
p3 <- mc.res60$plot
p4 <- komp.res20$plot
p5 <- komp.res40$plot
p6 <- komp.res60$plot
p7 <- komp.info.res20$plot
p8 <- komp.info.res40$plot
p9 <- komp.info.res60$plot

combined_plot <- arrangeGrob(p1, p2, p3, p4, p5, p6, p7, p8, p9, ncol = 3)

grid::grid.draw(combined_plot)
ggsave("docs/figure/figures.Rmd/kompute_test_sim_plot_v16.png", combined_plot, height=12, width=13)
```


